\subsection{Graph-Based Programming Models}

Many programming systems have designed for writing graph-based programs.  Some
good examples are the Dryad, Pregel, GraphLab, Ligra, Grace, Galois and
SociaLite systems.

The Dryad system~\cite{Isard:2007:DDD:1272996.1273005} combines computational
vertices with communication channels (edges) to form a data-flow graph. The
program is scheduled to run on multiple processors or cores and data is
partitioned during runtime. Routines that run on computational vertices are
sequential, with no synchronization. Dryad is better suited for a wider range of
problem domains than LM, however Dryad is not a language but a framework upon
which languages such as LM could be built.

The Pregel system~\cite{Malewicz:2010:PSL:1807167.1807184} is also graph based,
although programs have a more strict structure. They must be represented as a
sequence of iterations where each iteration is composed of computation and
message passing. Pregel is especially suited for large graphs since it aims to
scale to large architectures. LM does not impose iteration restrictions on
programs, however it is not as well suited for large graphs as Pregel.

GraphLab~\cite{GraphLab2010} is a C++ framework for developing parallel machine
learning algorithms. While Pregel uses message passing, GraphLab allows nodes to
have read/write access to different scopes through different concurrent access
models in order to balance performance and data consistency. While some programs
only require access to the local node's data, others may need to update edge
information. Each consistency model will provide different guarantees that are
better adapted to some algorithms. GraphLab also provides different schedulers
that dictate the order in which nodes are computed. The downside of GraphLab is
that GraphLab has a steep learning curve and programs must be written in C++,
requiring a lot of boilerplate code. LM programs tend to be smaller and easier to
reason about since they are written at a higher abstraction level.

Ligra~\cite{Shun:2013:LLG:2517327.2442530} is a lightweight framework for large
scale graph processing on a single multicore machine. Ligra exploits the fact
that most huge graph datasets available today can be made to fit in the main
memory of commodity servers. Ligra is a simple framework that exposes two main
interfaces: \texttt{EdgeMap} and \texttt{VertexMap}. The former applies a
function to a subset of edges of the graph, while the latter applies a function
to a subset of vertices. The functions passed as arguments are applied to either
a single edge or a single vertex and the user must ensure that the function can
be executed in parallel. The framework allows the use of
\emph{Compare-and-Swap}~(CAS) instructions when implementing functions in order
to avoid race conditions.

Grace~\cite{wang:asynchronous} is another graph-based framework for multicore
machines. Unlike Ligra, Grace programs are implemented from the point of view of
a vertex. Each vertex and edge can be customized with different data depending
on the target application. By default, programs are executed iteratively: for
each iteration the vertex program reads incoming edge messages, performs
computation and sends messages to the outbound edges. Since iterative programs
require synchronization after each iteration, Grace allows the user to relax
these constraints and implement customizable execution policies by implementing
code for describing which vertices are allowed to run and in which order. The
order is dictated by assigning a \emph{scheduling priority value}.

Galois~\cite{Pingali:2011:TPA:1993316.1993501} is a parallel programming model
with irregular applications based on graphs, trees and sets. A Galois parallel
algorithm is viewed as a parallel application of an \emph{operator} over an
irregular data structure which generate \emph{activities} on the data structure.
Such operator may, for instance, be applied to the node of the graph in order to
change its data or change the structure of its neighborhood, allowing for data
structure changes. In Galois, \emph{active elements} are nodes of the graph
where computation needs to be performed. Each operator performed on an active
element needs to acquire the \emph{neighborhood}, which are the nodes connected
to the active element that are involved in the operator's computation.
Furthermore, operators are selected for execution according to some specific
\emph{ordering}. From the point of view of the programmer, the active elements
are represented in a work-list, while operators can be implemented on top of
work-list's iterators. Galois supports speculative execution by allowing
operator rollback when an operator requires a node that is held by another
operator.

Ligra, Grace and Galois are not programming language but frameworks built on top
of other programming languages such as C and Java. Naturally, programs written
in these frameworks need to take into account several implementation details
such as \emph{compare-and-set} instructions, work-lists, and scheduling order.
LM programs are more abstract since reasoning is performed around logical rules
and logical facts which are much closer to the problem domain of graph-based
programs.

SociaLite~\cite{Seo:2013:DSD:2556549.2556572} is a Datalog-based language for
writing algorithms for social-network analysis. SociaLite takes Datalog programs
and compiles them to efficient distributed Java code. The language places some
restrictions on rules in order to make distribution possible, however, the
programmer is free to tell the system how to shard database's facts. Like LM,
these restrictions deal with the first argument of each fact, however, the first
argument is not related to a graph abstraction but is instead related to fact
distribution, allowing programmers to optimize how messages are sent between
machines.

\subsection{Sensor Network Programming Languages}

Many programming languages have been designed to help developers write programs
that run on sensor networks, which are networks that can be represented as
graphs.  Programming languages such as
Hood~\cite{Whitehouse:2004:HNA:990064.990079},
Tinydb~\cite{Madden:2005:TAQ:1061318.1061322} or
Regiment~\cite{Newton:2007:RMS:1236360.1236422} have been proposed, providing
support for data collection and aggregation over the network. These systems
assume that the network remains static and nodes stay in place.

Other languages such as Pleiades~\cite{Kothari:2007:REP:1250734.1250757},
LDP~\cite{4543691} or Proto~\cite{Beal:2006:IEE:1137236.1137354} go beyond
static networks and support dynamic reconfiguration. In Pleiades, the
programmer writes an application from the point of view of the whole
sensor network and the compiler transforms it into code that can be run on
each individual node.  LDP is a language derived from a method for
distributed debugging, that allows it to efficiently detect conditions on
variably-sized groups of nodes. It is based on the tick model, generating
a new set of condition matchers throughout the ensemble on each tick.
Like Pleiades and LDP, Proto also compiles global programs into locally
executed code.

Finally, the original Meld~\cite{ashley-rollman-iclp09} is also a language
designed for dynamic networks, namely, ensembles of robots. As we have seen
before, Meld is a logic programming language where programs are sets of logical
rules that infer over the state of the ensemble. Meld supports action facts and
sensing facts, which allow the robots to act on the world or sense the world,
respectively. Like Pleiades programs, Meld programs are also written from the
point of the view of the whole ensemble.

\subsection{Constraint Handling Rules}

Since LM is a bottom-up linear logic programming language, it also shares
similarities with Constraint Handling Rules
(CHR)~\cite{Betz:2005kx,Betz:2013:LBA:2422085.2422086}.  CHR is a concurrent
committed-choice constraint language used to write constraint solvers. A CHR
program is a set of rules and a set of constraints. Constraints can be consumed
or generated during the application of rules. Unlike LM, in CHR there is no
concept of rule priorities, but there is an extension to CHR that supports
them~\cite{DeKoninck:2007:URP:1273920.1273924}. Finally, there is also a CHR
extension that adds persistent constraints and it has been proven to be sound
and complete~\cite{DBLP:journals/corr/abs-1007-3829} in relation to the standard
formulation of CHR.

\subsection{Graph Transformation Systems}

Graph Transformation Systems (GTS)~\cite{Ehrig:2004vn}, commonly used to model
distributed systems, perform rewriting of graphs through a set of graph
productions. GTS also introduces concepts of concurrency, where it may be
possible to apply several transformations at the same time. In principle, it
should be possible to model LM programs as graph transformations: we directly
map the LM graph of nodes to GTS's initial graph and consider logical facts as
nodes that are connected to LM's nodes. Each LM rule is then a graph production
that manipulates the node's neighbors (the database) or sends new facts to other
nodes. On the other hand, it is also possible to embed GTS inside
CHR~\cite{Raiser:2011:AGT:1972935.1972938}.
