I would like to thank my advisors, for their help and encouragement during the
development of this thesis. I thank Prof. Ricardo Rocha for his help during
implementation and debugging, for his suggestions and for his immense attention
to detail. I thank Prof. Frank Pfenning for helping me understand proof theory
and for instilling a sense of rigor about my work. I thank Prof. Seth Goldstein
for his great insights and for giving me the encourage to pursue the ideas I
had during the last 4 years. I have certainly become a better researcher
because of you all.

Besides my advisors, I would like to express my gratitude to the following
persons. A special acknowledgment to Prof. Fernando Silva, for encouraging me
to apply to the PhD program and for believing that I could do it.  To Michael
Ashley-Rollman, for helping me understand his work on ensemble programming.
This thesis would not be possible without his initial contribution.

I am thankful to the Fundacao para a Ciencia e Tecnologia (FCT) for the
research grant SFRH/BD/51566/2011, to the Center for Research and Advanced
Computing Systems (CRACS), and to the Computer Science Department at CMU for
their financial support.

To my fellow friends from DCC-FCUP, specially Miguel Areias, Joao Santos, and
Joana Corte-Real, for their friendship and excellent work environment, which
certainly helped me during the time I spent in Porto.

To my roommate Salil Joshi, for his friendship during my time at CMU. To all my
other friends at CMU for their intellectually stimulating conversations.

My deepest gratitude for my parents and my sisters, for their support and for
teaching me the value of hard work and persistence. Finally, I would also like
to thank my wife Telma, for her affection and encouragement.

\begin{flushright}
Flavio Cruz \\
\end{flushright}
