We now describe the linear logic fragment used as a basis for LM.  Note that in
this thesis we follow the intuitionistic approach and use the sequent
calculus~\cite{gen35} to specify the logic. Our initial sequent is written as
$\Psi; \seqx{\Gamma}{\Delta}{C}$ and can be read as "assuming persistent
resources $\Gamma$ and linear resources $\Delta$ then $C$ is true".  More
specifically, $\Psi$ is the typing context which contains unique variables,
$\Gamma$ is a multi-set of persistent resources, $\Delta$ is a multi-set of
linear resources while $C$ is the proposition we want to prove.

We first have the \emph{simultaneous conjunction} $A \otimes B$ that packages
linear resources together. In the right rule, $A \otimes B$ is true if both $A$
and $B$ are true, and, in the left rule, it is possible to split $A \otimes B$
apart.

\input{appendix/fragment_otimes}

Next, we have the \emph{additive conjunction} $A \with B$ that allows us to
select between $A$ or $B$. In the right rule we must prove $A$ and $B$ using
the same resources, while in the left rule, we can select one of the
resources.

\input{appendix/fragment_with}

To express inference, we introduce the \emph{linear implication} connective
written as $A \lolli B$. For the right rule, we prove $A \lolli B$ by assuming
$A$ and then proving $B$, while in the left rule, we obtain $B$ by using some
linear resources to prove $A$.

\input{appendix/fragment_lolli}

Next, we introduce persistent resources written as $\bang A$. For the right
rule, we prove $\bang A$ by proving it without any linear resources. Likewise,
to use a persistent resource, we simply drop the $
\bang$. There is also a $\m{copy}$ rule that moves persistent resources from
$\Gamma$ to $\Delta$. Remember that $\Gamma$ contains persistent resources.

\input{appendix/fragment_bang}

Another useful connective is the \emph{multiplicative unit} of the $\otimes$
connective. It is written as $\one$ and is best understood as something that
does not need any resource to be proven.

\input{appendix/fragment_one}

Next, we introduce the \emph{quantification} connectives, namely \emph{universal
quantification} $\forall_{n:\tau}. A$ and \emph{existencial quantification}
$\exists_{n:\tau}. A$. These connectives use the typing context $\Psi$ because
they can introduce and read term variables from the context. The right and left duals of
those two connectives are dual.

\input{appendix/fragment_quant}

The judgment $\Psi \vdash M : \tau$ introduces a new term $M$ with type $\tau$
that does not depend on $\Gamma$ or $\Delta$ but may depend on the variables in
$\Psi$. In rules $\forall R$ and $\exists L$, the new $m$ variable introduced in
$\Psi$ must always be \emph{fresh}. We complete the linear logic system with the
\emph{cut rules} and the \emph{identity rule}:

\input{appendix/fragment_cut}
\input{appendix/fragment_id}
