In the Heat Transfer~(HT) algorithm, we have a graph where heat values are
exchanged between nodes. The program stops when the new heat values $H_i$
differ only a small $\epsilon$ from the old values $H_{i-1}$, where $\delta =
|H_i - H_{i-1}| \le \epsilon$. The algorithm works asynchronously, i.e., heat
values are updated using information as it arrives from neighboring nodes. This
increases concurrency since nodes do not need to synchronize between
iterations.

Figure~\ref{code:coord:ht} shows the HT rules that send new heat values to
neighbor nodes. In the first rule we added \code{add-priority} action fact to
increase the priority of the neighbor nodes for the case when the current node
has a large $\delta$. The idea is to prioritize the computation of heat values
of nodes (using \code{update}) that have a neighbor that changed significantly.
Multiple \code{add-priority} facts will increase the priority of a node so that
nodes with multiple large deltas will have more priority.

\begin{figure}[h!]
\begin{Verbatim}[numbers=left,fontsize=\codesize,commandchars=\\\[\]]
new-heat(A, New, Old),
Delta = fabs(New - Old),
Delta > epsilon
   -o {B | !edge(A, B) -o
         new-neighbor-heat(B, A, New),
         update(B), \underline[add-priority(B, Delta)]}.

new-heat(A, New, Old)
fabs(New - Old) <= epsilon
   -o {B | !edge(A, B) -o
         new-neighbor-heat(B, A, New)}.
\end{Verbatim}
  \caption{Coordination code for the Heat Transfer program.}
  \label{code:coord:ht}
\end{figure}

\iffalse
Fig.~\ref{results:ht} presents the scalability results for the regular
and coordinated version. The dataset used is a square grid with an inner square
with high heat nodes. Comparing the coordinated version with the regular
version, with 1 thread there is a 50\% reduction in run time, while
for 16 threads there is, on average, a 25\% reduction.
\fi

To further improve locality, we can split the second rule to avoid sending small
$\delta$ values if the target node is in another thread
(Fig.~\ref{code:coord:ht_better}). We use \code{thread-id} to retrieve the
thread \code{T} of the node \code{A} and match the \code{thread-id} of each
neighbor \code{B} against \code{T}. The comprehension in
lines~\ref{line:coord:ht_better_comp1}-\ref{line:coord:ht_better_comp2} only
generates \code{new-neighbor-heat} facts if \code{B} is in the same thread.

\iffalse
The \textbf{C/Local} line in
Fig.~\ref{results:ht} shows that this performs best.  However, this comes at
the price of increased errors in the computed heat values.
\fi

\begin{figure}[h!]
\begin{Verbatim}[numbers=left,fontsize=\codesize,commandchars=\\\[\]]
new-heat(A, New, Old)
fabs(New - Old) <= epsilon,
\underline[thread-id(A, T)]
   -o {B, T | !edge(A, B), \underline[thread-id(B, T)] |\label[line:coord:ht_better_comp1]
         new-neighbor-heat(B, A, New), \underline[thread-id(B, T)]},\label[line:coord:ht_better_comp2]
      \underline[thread-id(A, T)].
\end{Verbatim}

  \caption{To improve locality, we add an extra constraint to the second rule to
     avoid sending small $\delta$ values if the target node is in another thread.}
  \label{code:coord:ht_better}
\end{figure}

