To further understand how coordination facts work, we present other programs that
take advantage of them.

\subsection{MiniMax}\label{section:coord:minimax}
\input{coordination/minimax}
\subsection{N-Queens}\label{section:coord:nqueens}
\input{coordination/queens}
\subsection{Heat Transfer}\label{section:coord:ht}
\input{coordination/ht}

\subsection{Remarks}

We showed four applications that make use of the new coordination facilities of
LM. Coordination makes it possible for programmers to fine tune, if necessary,
the scheduling and load balancing of declarative programs in the same way that
regular LM computation is accomplished.

In the Single Source Shortest Path program, we demonstrated that coordination
makes it easy to avoid redundant computations and thus significantly reduce the
run time of several, relatively large, datasets.

In the MiniMax program, we used default priorities to minimize the memory usage
of the program and allow the declarative program to be far more competitive
with the corresponding program implemented in C++.

In the N-Queens program, we used partitioning facts to pin columns of the chess
board to threads and then improve the scalability of the program for
configurations where the number of columns is equal to the number of threads.
The partitioning mechanism allows the programmer to improve the locality of the
program by deciding where nodes and computation should be placed during
execution.

Finally, in the Heat Transfer program, we made use of both scheduling and
partitioning facts to improve performance and reduce inter-thread communication.
The sensing facts for partitioning makes it trivial to selectively do
computation by reasoning about the state of the underlying parallel system.
